Classless Java

### 2016-03-20 OOPSLA16/sections/Introduction.tex
Polishing

We present here a third alternative: the concept of \textbf{Interface
  based object oriented languages} (IB), where objects implements
interfaces, and they own the implementation for the behaviour, which
is structurally defined in their interface.

The literature provide good examples on how easy and modular can be
combining multiple sources of pure behaviour (traits~\cite{}), and how hard
it is to modularly combine multiple sources of behaviour \textbf{and}
state (multiple inheritance~\cite{}).\bruno{needs strenghtning}

@@

Our idea could be explained by defining a novel language, with new
syntax and semantic.  However, this would have a steep learning curve;
for the sake of providing a more accessible explanation, we will
encode \footnote{ In the same sense, you can encode prototype based
  programming over class based with the strategy pattern, and class
  based programming with prototypes by creating ``class'' objects.  }
our ideas over Java, relying on the new features of Java 8; namely
interface \emph{static methods}, and \emph{default methods}, allowing
interfaces to have method implementations. While the motivation for
default methods was to allow interfaces to be extended over time while
preserving backwards compatibility (\emph{interface evolution}),
default methods can also emulate \emph{traits}~\cite{scharli03traits}.
\footnote{The original notion of traits by Scharli et al. prescribes,
  among other things, that: 1) a trait provides a set of methods that
  implement behavior; and 2) a trait does not specify any state
  variables, so the methods provided by traits do not access state
  variables directly. Java 8 interfaces follow similar principles
  too. Indeed, a detailed description of how to emulate trait-oriented
  programming in Java 8 can be found in the work by Bono et
  al.~\cite{bono14}. The Java 8 team designing default methods, was
  also fully aware of that secondary use of interfaces, but it was not
  their objective to model traits: ``The key goal of adding default
  methods to Java was "interface evolution", not "poor man's
  traits"''~\cite{goetz13default}. As a result, they were happy to
  support the secondary use of interfaces with default methods as long
  as it did not make the implementation and language more complex.}
\marcoT{We could use Scala Traits to encode IB, but the simpler Java8
  model is a better starting point.}  \footnote{ \marco{may go in
    related work} Scala traits can declare \Q@def, val, var@ and
  \Q@lazy val@ However, while \Q@def@s can be overridden by
  \Q@val@/\Q@var@s, \Q@val@/\Q@var@s can not be overridden.\marco{we
    need to check this} A\Q@val@/\Q@var@ member is predestined to be
  instance state, and this is in sharp contrast with our model where
  the state is only a private detail of instances, and do not
  contaminate interface/trait/classes.}  \footnote{Scala traits have
  no constructor, but \Q@var@s and (\Q@lazy@) \Q@val@s can be
  ``pre-initialized''.}

@@

In the context of Java, what we propose is essentially a programming
style, where we never use classes\footnote{More precisely, we never
  use the \Q@class@ keyword.}.  We call this restricted version of
Java \emph{Class-less Java}.  Our proposed style enables powerful
object-oriented idioms, similar to (stateful) multiple inheritance and
field type refinement (allowing a kind of covariant setters
refinement).

As for CB, in IB programmers do not define objects directly, but
delegate the task to \emph{object interfaces}, whose role is similar
to non abstract classes in CB.

To model state, the idea is that all abstract operations in an Object
interface will be interpreted as \emph{abstract state operations}; The
abstract state operations include various common utility methods (such
as getters and setters, or clone-like methods). In the presence of
subtyping, such operations often require special care, as their types
need to be refined. Object interfaces provide support for
type-refinement and can automatically produce code that deals with
type-refinement adequately, while Java (and Scala) require explicit
type-refinement.

@@

Using Java annotation processors, we produce an implementation of
Classless Java, which allow us to stick to pure Java8. By annotating
with \Q|@Obj| the interfaces that represent object interfaces, we can
generate boring and repetitive code for interface instantiation and
type refinement.  Such code should not be needed in the first place in
a real IB language.

@@

We formally define the behaviour of our \Q|Obj| annotation and we provide a
prototype implementation using Java annotations to do AST
rewriting, allowing existing Java tools (such as IDEs) to work
out-of-the-box with our implementation. As a result, we
experiment object interfaces with several interesting Java programs,
and conduct various case studies.

>>>>

We present here a third alternative: the concept of \textbf{Interface
  based object oriented languages} (IB), where objects implement
interfaces, and they own the implementation for the behaviour, which
is structurally defined in their interface. In IB programmers do not define objects directly, but
delegate the task to \emph{object interfaces}, whose role is similar
to non abstract classes in CB languages. Key motivations for IB
languages are improved \emph{modularity} and \emph{multiple inheritance}~\bruno{I'm using multiple
  inheritance in a more abstract way than you Marco.}  The literature
provides good examples on how easy and modular it is to combine
multiple sources of pure behaviour, using mechanisms such as
traits~\cite{scharli03traits}. In Java multiple \emph{interface} inheritance has been
supported since inception, and in Java8 default methods~\cite{} bring some of
the advantages of traits into Java. In contrast, the literature~\cite{} is
also rich on how hard it is to modularly combine multiple sources of
behaviour \textbf{and} state with multiple \emph{implementation}
inheritance of classes.

@@

%\footnote{ In the same sense,
%  you can encode prototype based programming over class based with the
%  strategy pattern, and class based programming with prototypes by
%  creating ``class'' objects.}

IB could be explained by defining a novel language, with new syntax
and semantics. However, this would have a steep learning curve.  We
take a different approach instead. For the sake of providing a more
accessible explanation, we will embed our ideas directly into Java. 
Our IB embedding relies on the
new features of Java 8 for interfaces: interface \emph{static methods}; and
\emph{default methods}, which allow interfaces to have method
implementations.

@@

In the context of Java, what we propose is essentially a programming
style, where we never use classes\footnote{More precisely, we never
  use the \Q@class@ keyword.}.  We call this restricted version of
Java \emph{Classless Java}.  
Our proposed style enables powerful
object-oriented idioms, similar to (stateful) multiple inheritance and
field type refinement (allowing a kind of covariant setters
refinement). \bruno{needs strenghtening}

A key challenge lies in how to model state, which is fundamental to
have stateful objects. In IB and Classless Java, the idea is that all abstract operations in an Object
interface will be interpreted as \emph{abstract state operations}. The
abstract state operations include various common utility methods (such
as getters and setters, or clone-like methods). In the presence of
subtyping, such operations often require special care, as their types
need to be refined. Object interfaces provide support for
type-refinement and can automatically produce code that deals with
type-refinement adequately, while Java (and Scala) require explicit
type-refinement.\bruno{strenghtening needed!}

@@

Using Java annotation processors, we produce an implementation of
Classless Java, which allow us to stick to pure Java8. By annotating
with \Q|@Obj| the interfaces that represent object interfaces, we can
generate boring and repetitive code for interface instantiation and
type refinement. Such code should not be needed in the first place in
a real IB language, and the annotation processor allows us to
transparently hide it from Java programmers. 
The implementation works by performing \emph{on-the-fly} AST
rewriting, allowing existing Java tools (such as IDEs) to work
out-of-the-box with our implementation. Moreover, the 
implementation blends Java's conventional CB style and IB smoothly. 
As a result, we experiment object interfaces with several interesting Java programs,
and conduct various case studies. 
Finally, we also formally define the behaviour of our \Q|Obj| implementation.

@@

### 2016-03-20 OOPSLA16/sections/RelatedWork.tex
Polishing

>>>>

\bruno{next comes some text that may be handy for related work}

While the motivation for default methods was to allow
interfaces to be extended over time while preserving backwards
compatibility (\emph{interface evolution}), default methods can also
emulate \emph{traits}~\cite{scharli03traits}.  \footnote{The original
  notion of traits by Scharli et al. prescribes, among other things,
  that: 1) a trait provides a set of methods that implement behavior;
  and 2) a trait does not specify any state variables, so the methods
  provided by traits do not access state variables directly. Java 8
  interfaces follow similar principles too. Indeed, a detailed
  description of how to emulate trait-oriented programming in Java 8
  can be found in the work by Bono et al.~\cite{bono14}. The Java 8
  team designing default methods, was also fully aware of that
  secondary use of interfaces, but it was not their objective to model
  traits: ``The key goal of adding default methods to Java was
  "interface evolution", not "poor man's
  traits"''~\cite{goetz13default}. As a result, they were happy to
  support the secondary use of interfaces with default methods as long
  as it did not make the implementation and language more complex.}
\marcoT{We could use Scala Traits to encode IB, but the simpler Java8
  model is a better starting point.}  \footnote{ \marco{may go in
    related work} Scala traits can declare \Q@def, val, var@ and
  \Q@lazy val@ However, while \Q@def@s can be overridden by
  \Q@val@/\Q@var@s, \Q@val@/\Q@var@s can not be overridden.\marco{we
    need to check this} A\Q@val@/\Q@var@ member is predestined to be
  instance state, and this is in sharp contrast with our model where
  the state is only a private detail of instances, and do not
  contaminate interface/trait/classes.}  \footnote{Scala traits have
  no constructor, but \Q@var@s and (\Q@lazy@) \Q@val@s can be
  ``pre-initialized''.}
